\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

\begin{tabular}{@{} >{\bfseries}p{0.2\textwidth} p{0.77\textwidth} @{}}
\texorpdfstring{\MakeUppercase{Keywords}}{Keywords} &
telecom tower sites; hybrid energy systems; solar PV; battery storage;
diesel logistics; reinforcement learning; constrained Markov decision process;
inventory management; action masking; expected energy not served;
model predictive control
\end{tabular}
\par
\begin{doublespacing}

Reliable and low-carbon energy management for hybrid telecom tower sites
remains a critical challenge in regions with weak or intermittent electricity
supply. Conventional rule-based controllers and offline optimisation tools
such as HOMER Pro rely on fixed thresholds and deterministic assumptions,
making them unable to adapt to stochastic solar generation, grid outages, and
fuel logistics uncertainty.

This thesis studies the operational control of hybrid
solar--battery--diesel telecom tower systems under uncertainty in renewable
generation, grid availability, and diesel delivery. The problem is formulated
as a Constrained Markov Decision Process (CMDP) that explicitly models
battery state-of-charge limits and diesel inventory dynamics, including diesel replenishment ordering and stochastic delivery lead times.

A data-grounded simulation environment is developed using the ITU~2024
Smart Energy Scheduling dataset, comprising operational traces from ten
telecom base station sites at hourly resolution. Reinforcement learning
policies are trained using a Maskable Proximal Policy Optimisation
(\texttt{MaskablePPO}) agent with hard action masking to enforce operational
feasibility constraints at every timestep.

The proposed inventory-aware reinforcement learning policy (RLInv) is
evaluated against eight comparator policies --- rule-based, reinforcement
learning, and optimisation-based, including a receding-horizon Model Predictive
Control (MPC) baseline and a perfect-information Oracle-MPC reference ---
across ten sites, ten independent training seeds, and four delivery-logistics
scenarios of increasing severity, using paired seed-level statistical
comparisons throughout. Across the four representative logistics scenarios,
RLInv achieves the lowest mean Expected Energy Not Served (EENS) in three
scenarios and remains statistically competitive in the fourth, and this
advantage is not attributable to superior dispatch or
forecasting alone: the gap to Oracle-MPC, which assumes perfect
future information, does not shrink under increasing logistics
stress. Isolating the contribution of jointly learned replenishment from a
cleaner ablation shows a benefit that is scenario-dependent rather than
universal --- significant under normal and severe logistics stress, but not
distinguishable from a classical ordering rule under moderate (monsoon-scenario)
stress. Action masking is shown to improve dispatch-policy robustness
specifically under logistics conditions harsher than the training
distribution, rather than acting as the primary safety mechanism at
deployment time.

These results indicate that the value of jointly learning replenishment
and dispatch is conditional on the logistics regime rather than
unconditional: reinforcement learning is not shown to be universally
superior to classical optimisation, but joint optimisation of
replenishment and dispatch is shown to provide measurable value
specifically under uncertainty that classical rule-based coordination
cannot easily represent. This conditional finding, together with a
systematically verified evaluation framework spanning primary comparison,
mechanism ablations, and multiple robustness checks, is offered as a more
useful basis for deployment decisions than an unconditional claim of RL
superiority would have been.

\end{doublespacing}
