\chapter{Discussion and Conclusion}
\label{chap:conclusion}
\setlength{\emergencystretch}{1.5em}

Chapter~\ref{chap:results} reported what the experimental evidence
shows. This chapter discusses what it means: which design choices the
evidence actually supports, why the scenario-dependent results
occurred rather than the uniformly positive result originally
hypothesised, what this thesis contributes, its limitations, and the
research directions that follow.

% ============================================================
\section{Summary of the Work}
\label{sec:conclusion_summary}
% ============================================================

This thesis investigated whether a unified reinforcement learning
agent, jointly optimising energy dispatch and diesel inventory
ordering, improves the operational reliability of off-grid and
weak-grid telecom base stations relative to decoupled classical and
optimisation-based approaches. The gap motivating this work is
practical: prior RL-based energy management for remote base stations
routinely models battery dispatch and generator scheduling but omits
the upstream replenishment decision --- when and how much diesel to
order --- despite this being a primary driver of service interruption
in fuel-constrained deployments.

The problem was framed as a constrained Markov decision process in
which a single policy (RLInv) controls generator commitment and diesel
order quantity jointly, with battery dispatch handled by a fixed
priority rule and an 11-dimensional observation vector including
explicit inventory and pipeline state (Chapter~\ref{chap:problem}).
RLInv was evaluated against eight comparator policies spanning
rule-based, reinforcement learning, and optimisation-based approaches ---
including, notably, a receding-horizon MPC baseline and an
Oracle-MPC optimistic dispatch reference under perfect future load, solar-generation and grid-availability information, added specifically to distinguish whether any
observed advantage reflects better \emph{ordering} or merely better
\emph{dispatch} (Chapter~\ref{chap:methodology}). The final campaign
evaluated all nine core policies across 10 ITU/Zindi sites, 10
independent training seeds, and four delivery-logistics scenarios of
increasing severity (Normal, Delayed, Monsoon, Extreme), with paired
site-seed statistical comparisons for the primary policy and ablation analyses
(Chapter~\ref{chap:experimental-execution}).

% ============================================================
\section{Key Results and Discussion}
\label{sec:conclusion_results}
% ============================================================

\subsection{What the primary comparison shows}

Across the evaluated scenarios, RLInv consistently ranks among the
best-performing policies and achieves the lowest mean EENS in three of
the four representative logistics scenarios (Normal, Delayed, Extreme),
remaining competitive with the best classical baseline under Monsoon.
Critically, this advantage is not explained by superior
dispatch alone: a rolling-horizon MPC baseline with a classical
ordering rule performs no better than the simplest rule-based ordering
policy, and the gap to Oracle-MPC --- which has access to
realised future load, solar generation and grid availability for dispatch planning --- is largest under
Monsoon but narrows again under Extreme, where physical fuel
availability increasingly constrains all policies. Taken together, these results show that dispatch quality and
forecasting alone do not explain RLInv's overall performance. The
ablation results indicate that jointly learned replenishment
contributes under some logistics regimes, although its independent
effect is not uniform across scenarios. The advantage is also not achieved through
disproportionate fuel use: RLInv consumes only 3.6\% more diesel than
the best classical baseline (205.7 vs.\ 198.5\,kWh), consistent with a
genuine operational trade-off rather than an inflated comparison.

Read together, the Oracle-MPC comparison and the mechanism results
that follow point to a single underlying message. RLInv's contribution
is not that reinforcement learning is universally better than classical
optimisation --- Oracle-MPC, MPC, and the classical $(s,S)$ rule remain
strong references throughout. It is that joint optimisation of
replenishment and dispatch becomes valuable specifically under
uncertainty that classical rule-based coordination cannot easily
represent, and that this value is conditional on the logistics regime
rather than unconditional.

\subsection{Why the mechanism results are scenario-dependent, not uniform}

The three pre-registered hypotheses were tested through paired,
seed-level ablations rather than a single aggregate comparison, and the
honest answer in each case is more nuanced than a simple confirmed/rejected
verdict.

\textbf{Joint replenishment and dispatch learning (H1).} The cleaner
practical test (RLInv vs.\ A6) replaces learned ordering with an
external calibrated $(s,S)$ rule and retrains dispatch using a reduced
two-action space; it therefore reduces the confounding present in
RLInv--TrackB, but does not hold the dispatch architecture fully
constant. This comparison shows a
significant RLInv advantage under Normal and Extreme logistics, no
difference under Delayed, and a result that nominally favours the
classical-ordering comparator under Monsoon. This is not the uniform
"joint learning always wins" result originally anticipated. Contrary to
the original hypothesis, the paired Phase 3 analysis shows that the
benefit of joint replenishment learning is scenario-dependent rather
than universal; this finding illustrates the value of large-scale
replication, which refined the original interpretation obtained from
the earlier pilot study. The results indicate that the additional
benefit from jointly learned replenishment depends on the operating
regime: it is measurable under Normal and Extreme logistics but not
under Delayed or Monsoon conditions.

\textbf{Action masking (H2).} A6 substantially outperforms TrackB
under Monsoon and Extreme logistics. However, the comparison differs
in both masking and observation design (Chapter~\ref{chap:methodology}), so the gap cannot be
attributed to masking alone. The cleaner RLInv--A7 comparison, which
retains the same observation and joint-control architecture, finds no
statistically distinguishable mean-EENS effect from removing masking.
The evidence therefore does not establish a universal independent
performance benefit from masking, although masking remains the
mechanism used to enforce action feasibility. The observed A6--TrackB
gap is architecture-dependent and appears only under stressed
logistics.

\textbf{Inventory observability (H3).} RLInv achieves numerically lower
EENS than an ablation with inventory and ordering-history observations
removed, in every scenario, but the difference never reaches
significance (closest at Extreme, $p=0.058$). This is reported as a
genuine finding, not a shortfall: it suggests the agent can infer much
of the required information indirectly from other observable state
(battery state-of-charge, load, solar generation, grid availability,
and time-of-day features). The combined contribution of the five
removed logistics-related channels is modest under the final protocol;
because they are removed together, the experiment does not isolate the
contribution of inventory level from pending-order and
ordering-history information.

\subsection{Behaviour, robustness, and the exploratory extension}

Behavioural analysis shows RLInv's ordering frequency decreasing
monotonically from 2.16 orders per episode under Normal logistics to
1.34 under Extreme, consistent with the policy reducing new orders as
delivery completion within the evaluation horizon becomes increasingly
unlikely, while stockout events remain at zero even under Extreme
stress --- non-zero EENS occurs without recorded diesel stockout or
hard constraint violation. A sensitivity analysis across a wider
sequence of mean delivery
lead times shows RLInv's advantage over the best classical baseline
peaking at 29.5\,kWh around 72\,h and narrowing to 8.6\,kWh by 336\,h;
this narrowing reflects a physical constraint common to every policy
(tank capacity bounds how much can be pre-ordered regardless of
strategy), not RLInv ceasing to function.

The primary comparison is robust to reward-coefficient choice (EENS
varies by only 3.32\,kWh across a 20-fold range of the unmet-load
penalty), to tank capacity (RLInv's advantage narrows as tank capacity
increases, showing that sufficiently large physical buffers reduce the
value of proactive learned ordering), and to the parametric form
assumed for delivery delay (policy rankings are preserved under
lognormal and Weibull alternatives, with RLInv less sensitive to this
modelling choice than the classical baseline). One genuine limitation
was found under stochastic (rather than deterministic) grid outages: a
policy trained under this harder condition improves under Normal
logistics but degrades under Delayed, Monsoon, and Extreme, most
severely at Monsoon (+43.4\%). This pattern suggests a possible
trade-off between adaptation to grid uncertainty and fuel conservation
under prolonged delivery
disruption, reported as an identified limitation rather than resolved.

An exploratory extension (RLInv-DA) investigated whether explicit
delay-history tracking, motivated by Phase-I reviewer feedback,
improves on RLInv further. Despite a verified, working mechanism (a
correctly updating delay estimate under a persistent two-regime
supplier model), the extension produced no consistent EENS improvement
across the full evaluation grid on the narrower evaluation grid used
for this extension. This may indicate that the existing state already
contains sufficient indirect information for the tested conditions, or
that the simple EWMA signal does not capture the relevant supplier
dynamics; the added complexity is not justified at the scale evaluated.

% ============================================================
\section{Thesis Contributions}
\label{sec:conclusion_contributions}
% ============================================================

This thesis makes the following contributions to RL-based energy
management of remote telecom infrastructure:

\begin{enumerate}

  \item \textbf{A CMDP formulation with explicit, jointly controlled
  inventory and dispatch decisions.} The inclusion of diesel inventory
  and in-transit pipeline state alongside battery state-of-charge and
  grid availability enables anticipatory ordering that reacts to
  lead-time uncertainty before a stockout occurs, evaluated against
  eight comparator architectures rather than assumed superior.

  \item \textbf{An empirical comparison of coupled and decomposed
  control.} The study compares RLInv with rule-based, RL-based, and
  optimisation-based alternatives. The A6 ablation provides a cleaner
  practical assessment of learned replenishment, while its reduced
  dispatch action space is acknowledged as a remaining design
  difference. MPC and Oracle-MPC help contextualise dispatch and
  forecasting quality; they do not by themselves isolate ordering. The
  result is a genuinely
  scenario-dependent finding, reported honestly rather than as a single
  uniform verdict.

  \item \textbf{A qualified assessment of action masking.} Action
  masking is implemented as deterministic feasibility filtering. The
  cleaner RLInv--A7 comparison finds no statistically distinguishable
  mean-EENS effect when ordering is jointly learned, while the larger
  A6--TrackB gap under stressed logistics is confounded by differences
  in observation design. This distinction prevents feasibility
  enforcement from being mistaken for a universal performance gain.

  \item \textbf{A reproducible multi-stage evaluation framework.} The
  primary comparison and core ablations use the common 10-site,
  10-seed, and four-scenario protocol. Additional reward, tank,
  distributional, stochastic-grid, and delay-aware studies use
  explicitly documented experimental grids appropriate to their
  respective questions. Canonical source files and statistical units
  are stated for each analysis, with every figure and
  statistic independently re-derived from source during this work
  rather than assumed correct.

  \item \textbf{An honestly reported exploratory extension.} The
  delay-aware extension (RLInv-DA) is reported as a verified
  exploratory extension that did not produce a consistent improvement,
  rather than omitted.

\end{enumerate}

% ============================================================
\section{Critique and Limitations}
\label{sec:conclusion_limitations}
% ============================================================

The following limitations define the scope within which these results
should be interpreted.

\begin{description}

  \item[Synthetic delivery-pipeline state.] The diesel-delivery process
  and in-transit pipeline variables are synthetic overlays because the
  ITU/Zindi dataset contains no order or delivery records. Load, PV
  generation, and grid availability are dataset-derived, while
  equipment parameters and operating constraints follow the
  engineering assumptions documented in Chapters~\ref{chap:problem}
  and~\ref{chap:methodology}.
  Its geometric (and, as a robustness check, lognormal and Weibull)
  parameterisation is a modelling choice, not an empirical distribution
  fitted to real vendor data.

  \item[Discrete, tank-relative order quantities.] Order quantities are
  restricted to three levels expressed as a percentage of tank capacity
  (0\%, 30\%, 60\%), rather than continuous or site-specific fixed
  litre amounts. This simplifies the action space but means the
  precise quantity granularity available to a real operator is not
  represented.

  \item[Single-tower scope.] Each site is evaluated as an independent
  control problem. Multi-tower coordination --- for example, several
  towers competing for delivery slots from a shared supplier during a
  regional disruption --- is outside the scope of the CMDP formulated
  here.

  \item[Stochastic grid outage trade-off unresolved.] Training under a
  stochastic rather than deterministic grid-outage model improves
  Normal-scenario performance but degrades performance under logistics
  stress; this trade-off is documented (Chapter~\ref{chap:results}) but
  not resolved within this thesis.

  \item[RLInv-DA evaluated on a narrower grid.] The delay-aware
  extension's evaluation uses three sites and three training seeds
  rather than the full ten-site, ten-seed grid used elsewhere, since it
  was scoped as an exploratory check rather than a primary result; its
  finding of no consistent improvement should be read with this narrower scope in mind.

  \item[No real vendor delivery data.] Every distributional-robustness
  check (geometric, lognormal, Weibull) is a comparison between
  parametric assumptions, not a validation against observed delivery
  logs, which do not exist in the available dataset.

\end{description}

% ============================================================
\section{Future Work}
\label{sec:conclusion_future}
% ============================================================

Several of the directions identified during the original project
proposal --- multi-scenario evaluation, an expanded seed count,
optimisation-based baselines, and evaluation across heterogeneous site
profiles --- have
since been completed and are reported in
Chapter~\ref{chap:results} rather than remaining open. The four
directions below are ordered by how directly they continue this
thesis's own findings, from richer logistics modelling through to
deployment generalisation; two smaller extensions follow at the end.

\textbf{1.\ Richer logistics and supplier-behaviour modelling.} Two
results in this thesis point the same way. First, RLInv-DA showed that
a simple online estimator (EWMA) of realised delivery delay does not
improve replenishment at the evaluation scale tested. Future work could
compare richer state-estimation approaches, including Bayesian filters,
state-space models, and learned delay predictors, using empirical
supplier data where available.
Second, training under stochastic grid outages improves Normal-scenario
performance at the cost of degraded performance under logistics stress;
a natural extension is to train jointly under stochastic grid
\emph{and} varied logistics scenarios simultaneously, testing whether
this trade-off is an artefact of modelling only one source of
uncertainty at a time. Both directions aim at the same underlying gap:
richer, more realistic supplier and infrastructure behaviour than the
geometric and two-state abstractions used here.

\textbf{2.\ Fleet-level multi-agent reinforcement learning.} Realistic
deployments often involve several towers sharing a common diesel
supplier and competing for delivery capacity during regional
disruptions. The single-tower CMDP formulated here cannot capture this
coordination problem; a multi-agent extension in which each tower is an
independent agent but lead times are coupled through shared logistics
capacity is the most direct route to operationally realistic deployment
at scale, and the natural fleet-level counterpart to the single-tower
joint-optimisation result established here.

\textbf{3.\ Validation against empirical delivery datasets.} Every
distributional-robustness check in this thesis (geometric, lognormal,
Weibull) compares parametric modelling assumptions against one another,
not against observed delivery logs, since none exist in the available
dataset. Partnering with a telecom operator or fuel logistics provider
to obtain real delivery-time records would allow the lead-time model
itself to be validated empirically, rather than only stress-tested
across plausible parametric forms.

\textbf{4.\ Transfer learning and cross-region deployment.} The
cross-site evaluation in this thesis trains and evaluates a separate
policy per site; it is a per-site robustness check, not a test of
whether a policy trained on a subset of sites transfers to unseen ones
or to a different deployment region. Training on a subset of sites and
evaluating on held-out sites, or on sites from a different geographic
region with different solar, load, and outage characteristics, would
test this directly and is the natural precursor to any claim of
practical, low-cost deployment across a large or geographically diverse
fleet.

Two smaller extensions follow directly from the objective function and
state representation already defined in this thesis, without requiring
new experimental infrastructure. \textbf{Carbon and multi-objective
reporting}: future work could extend the current energy- and
reliability-based objective with explicit emissions and
battery-degradation terms, and report EENS--cost--CO$_2$ Pareto
trade-offs rather than relying on a single scalar reward. \textbf{Solar forecast integration}: RLInv
currently observes only realised PV output; augmenting the observation
with a short-horizon solar forecast, available in practice from
numerical weather prediction services, would let the agent anticipate
low-PV periods rather than reacting to them, a natural complement to
the delay-aware extension already explored on the ordering side.

% ============================================================
\section{Closing Remarks}
\label{sec:conclusion_closing}
% ============================================================

This thesis set out to answer a specific question: does making the
diesel inventory pipeline visible to a reinforcement-learning
controller, and letting it decide when and how much to order,
measurably improve service reliability at fuel-constrained telecom
base stations, relative to both classical decomposition and
optimisation-based dispatch? The answer, from a carefully audited
evaluation across ten sites, ten seeds, and
four logistics regimes, is conditional. The additional
benefit from jointly learned replenishment is measurable under Normal
and Extreme logistics, but is not statistically distinguishable under
Delayed or Monsoon conditions.

These conditional results are operationally useful because a
practitioner deciding how to deploy energy-management software at a
remote site now has evidence about \emph{when} joint optimisation is
worth its added complexity, rather than a claim that it always is. The
decision to test three pre-registered hypotheses rather than assert
that reinforcement learning is unconditionally superior allowed the
scenario-dependent results (H1's Monsoon exception, the
architecture-dependent H2 findings, H3's non-significance, RLInv-DA's
finding of no consistent improvement) to
be reported as genuine evidence rather than suppressed in favour of a
cleaner narrative.

Energy security at remote telecom towers is not a niche engineering
problem. In many low- and middle-income countries, base station uptime
determines access to emergency communication, mobile banking, and
public health services, and diesel logistics account for a substantial
share of operating cost in weak-grid deployments. A controller that is
reliable under routine conditions and whose limits under logistics
stress are precisely characterised, rather than assumed away, is a
more trustworthy basis for real deployment decisions than one whose
advantages are reported without qualification.
