\chapter{Methodology and Implementation}
\label{chap:methodology}

Chapter~\ref{chap:problem} defined the formal CMDP governing telecom tower
energy management. This chapter presents the implemented solution
framework: the simulator and training pipeline (Sections~\ref{sec:overview}
and~\ref{sec:rationale}), the environment implementation
(Section~\ref{sec:env_impl}), the reward design and its justification
(Section~\ref{sec:reward_design}), the full comparison-policy set
(Section~\ref{sec:comparison_policies}), the experimental protocol
(Section~\ref{sec:evaluation_protocol}), and the deviations between the
formal model and what was actually built
(Section~\ref{sec:impl_deviations}).

% ============================================================
\section{Methodology Overview}
\label{sec:overview}
% ============================================================

Figure~\ref{fig:chapter_roadmap} orients the reader before the technical
detail that follows: the CMDP formulated in Chapter~\ref{chap:problem} is
solved by the RLInv agent designed in this chapter, trained and
evaluated as described here, with the resulting evidence reported in
Chapter~\ref{chap:results}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  node distance=0.9cm,
  box/.style={draw, rounded corners, align=center,
              minimum width=3.4cm, minimum height=0.7cm, font=\small}
]
\node[box] (problem)  {Problem (Ch.~\ref{chap:problem})};
\node[box, below of=problem]   (cmdp)    {CMDP Formulation};
\node[box, below of=cmdp]      (rlinv)   {RLInv Design};
\node[box, below of=rlinv]     (train)   {Training};
\node[box, below of=train]     (eval)    {Evaluation};
\node[box, below of=eval]      (stats)   {Statistics};
\node[box, below of=stats]     (results) {Results (Ch.~\ref{chap:results})};

\draw[->] (problem) -- (cmdp);
\draw[->] (cmdp)    -- (rlinv);
\draw[->] (rlinv)   -- (train);
\draw[->] (train)   -- (eval);
\draw[->] (eval)    -- (stats);
\draw[->] (stats)   -- (results);
\end{tikzpicture}
\caption{Chapter roadmap: from the formal problem to the reported
evidence.}
\label{fig:chapter_roadmap}
\end{figure}

Figure~\ref{fig:workflow} shows the overall pipeline from raw data to
operational metrics. The ITU/Zindi dataset is loaded and split per site
into a 45-day training block and a 15-day test block. Each site's time
series drives a Gymnasium-compatible \texttt{TelecomEnv} simulator.
\texttt{MaskablePPO} agents are trained within this simulator, producing a
learned policy that is subsequently frozen and evaluated across all ten
sites under four delivery-logistics scenarios. Operational
metrics---primarily Expected Energy Not Served (EENS)---are computed from
the evaluation rollouts. Exploratory data analysis of the dataset is
presented in Chapter~\ref{chap:datasets-setup}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  node distance=1.8cm,
  box/.style={draw, rounded corners, align=center,
              minimum width=4.0cm, minimum height=0.9cm, font=\small}
]
\node[box] (data)    {ITU/Zindi Dataset\\{\scriptsize 10 sites, 60 days each}};
\node[box, below of=data]   (split)   {Train / Test Split\\{\scriptsize 45 days train \textbar\ 15 days test}};
\node[box, below of=split]  (env)     {TelecomEnv Simulator\\{\scriptsize Gymnasium, hourly $\Delta t$}};
\node[box, below of=env]    (train)   {RL Training\\{\scriptsize MaskablePPO, 10 seeds}};
\node[box, below of=train]  (policy)  {Learned Policy\\{\scriptsize Frozen weights}};
\node[box, below of=policy] (eval)    {Evaluation\\{\scriptsize 10 sites, 4 scenarios, 10 episodes/seed}};
\node[box, below of=eval]   (metrics) {Operational Metrics\\{\scriptsize EENS, Diesel, Stockouts}};

\draw[->] (data)    -- (split);
\draw[->] (split)   -- (env);
\draw[->] (env)     -- (train);
\draw[->] (train)   -- (policy);
\draw[->] (policy)  -- (eval);
\draw[->] (eval)    -- (metrics);
\end{tikzpicture}
\caption{End-to-end workflow of the proposed reinforcement-learning
framework, from dataset ingestion through policy training to operational
evaluation.}
\label{fig:workflow}
\end{figure}


% ============================================================
\section{Design Decisions}
\label{sec:rationale}
% ============================================================

\subsection{Why PPO rather than value-based methods?}

The decision problem combines generator commitment and diesel ordering over
long horizons (720-step training episodes). Value-based methods such as DQN
enumerate $Q$-values for each action combination and have been reported to
exhibit
instability in long-horizon environments with stochastic
dynamics~\citep{mnih2015humanlevel}. Proximal Policy Optimisation (PPO) was
selected because its clipped
policy updates have been widely used for stable training in stochastic
control and energy-management
applications~\citep{schulman2017proximal,perera2021application,zhang2021deep}.
The choice also integrates directly with the action-masking implementation
used in this work.

\subsection{Why MaskablePPO rather than penalty-based constraint handling?}

Penalty-based constraint handling adds infeasibility costs to the reward.
A finite penalty cannot guarantee zero hard violations: if the coefficient
is poorly calibrated the policy still explores infeasible actions, and a
diesel stockout or battery deep-discharge causes an operational constraint
violation or loss of service. \texttt{MaskablePPO} from \texttt{sb3-contrib}
instead sets infeasible action logits to $-\infty$ before the softmax, making
the probability of any infeasible action exactly zero by construction. This
converts the constraint set $\mathcal{C}(s_t)$ into a hard structural
property of the policy distribution rather than something the policy must
learn to avoid; a violation-backstop coefficient ($\mu$) is still retained
(Section~\ref{sec:reward_design}) as a secondary safeguard. Ablation~A7
trains an otherwise identical agent without action masking, giving a direct
empirical test of this claim (Chapter~\ref{chap:results}).

\subsection{Why a unified agent (RLInv) alongside decomposed control?}

Classical decomposed approaches use separate controllers for inventory
ordering and energy dispatch. This ignores the bidirectional coupling: the
dispatch controller depletes inventory, while inventory level constrains
which dispatch actions are feasible. A unified agent can in principle
exploit this coupling by jointly optimising both decisions from a single
reward signal. Whether that benefit justifies the added learning complexity
is the central empirical question (Hypothesis~H1). Two decomposed
comparators test this: TrackB, where a calibrated $(s,S)$ rule handles
ordering and a separately trained PPO agent handles dispatch; and A6, a
cleaner variant that holds the dispatch architecture fixed and substitutes
only the ordering mechanism (Section~\ref{sec:comparison_policies}). All
architectures are trained and evaluated under identical conditions;
neither outcome is assumed in advance.

\subsection{Why a discrete action space?}

Although battery dispatch could be formulated as a continuous RL action,
rural charge-controllers follow fixed priority rules rather than
continuous analogue setpoints. The implemented system therefore delegates
battery dispatch to a deterministic priority rule, focusing RL learning
capacity on the logistics decisions (generator commitment and ordering)
where non-myopic planning is most valuable.

\subsection{Why a flat MLP rather than modular encoders?}

With an 11-dimensional observation vector, a two-layer flat MLP
$[256,\,256]$ provides sufficient representational capacity. Modular
domain-specific encoders introduce additional architectural
hyperparameters (encoder depth, output dimension, concatenation
structure) with no demonstrated benefit at this input dimensionality. The
flat design simplifies implementation and is directly supported by the
Stable-Baselines3 \texttt{MlpPolicy} without a custom feature extractor.


% ============================================================
\section{Environment Implementation}
\label{sec:env_impl}
% ============================================================

\subsection{Structure and observation}

The environment is implemented as a Gymnasium-compatible class
\texttt{TelecomEnv} operating at hourly resolution ($\Delta t = 1$\,h)
with 720-step training episodes (30 simulated days) and 360-step
evaluation episodes (15 days, matching the held-out test split exactly).
During training, a random start index is sampled within the 45-day
training block; evaluation is fixed to the 15-day test block. Episodes
terminate after the fixed step count; no early termination condition is
used.

The 11-dimensional observation at each step is:
\begin{multline}
s_t = \bigl[
  \SoC_t,\;
  \Inv_t,\;
  \texttt{pending\_flag}_t,\;
  \texttt{pending\_qty}_t,\;
  P_t^{\PV},\; \\
  P_t^{\Load},\;
  G_t,\;
  \sin(2\pi h_t/24),\;
  \cos(2\pi h_t/24),\;
  \tau_t,\;
  R_t
\bigr].
\end{multline}
The pending-delivery state uses a binary flag (order in transit?) and a
scalar quantity (fuel expected on arrival), which together are logically
equivalent to the single pipeline scalar $\Pipe_t$ of
Chapter~\ref{chap:problem} under the single-outstanding-order constraint.
Two further channels complete the 11-dimensional vector. \texttt{hours\_since\_order\_n}
($\tau_t$) is the elapsed time since the last order was placed,
resetting to zero whenever a new order is issued; it is active in the
standard experiments. \texttt{delivery\_remaining\_n} ($R_t$) is present
in the observation vector but zero in the principal geometric,
no-ETA campaign reported throughout this thesis; it exists for
compatibility with a separate lognormal-forecast extension outside the
primary scope of this work. RLInv-DA (Chapter~\ref{chap:results})
additionally adds an EWMA delay estimate as a twelfth channel, producing
a 12-dimensional observation for that exploratory extension only.
Hour-of-day is encoded as sine--cosine features so that hours~23 and~0 are
adjacent in feature space. Grid availability $G_t$ is treated as an
exogenous stochastic signal derived from the site time series rather than
a decision variable. All continuous features are normalised online using
\texttt{VecNormalize}; statistics are frozen before evaluation.

\subsection{Action space, battery rule, and inventory}

Diesel inventory is tracked in kWh-equivalent units. Tank capacity is
$C^{\Tank} = 72\,\bar{D}$ (approximately three days of full-load backup,
where $\bar{D}$ is the site mean hourly load). A generator start is
permitted only if
$\Inv_t \geq 1.20 \times P_{\mathrm{rated}}^{\DG} \times 2\,\mathrm{h}$,
ensuring sufficient fuel for at least two hours of operation with a 20\%
safety margin.

The action space is discrete with six elements encoding
$\DG_t \in \{0,1\}$ and $Q_t \in \{0,\;0.30\,C^{\Tank},\;0.60\,C^{\Tank}\}$
--- order quantities are expressed as a percentage of tank capacity rather
than a fixed litre value, so that the same action encoding applies
consistently across sites of different sizes.

\begin{quote}
\textbf{Implementation Note.} This differs from the fixed-litre order
quantities described in Chapter~\ref{chap:problem}'s formal action space
(Equation~\ref{eq:action}); the tank-relative encoding used here is the
one actually implemented and evaluated throughout this thesis.
\end{quote}

Battery
dispatch is not a learnable action; it is determined by a fixed
energy-priority rule inside the simulator:
\[
\text{PV} \;\to\; \text{Grid} \;\to\; \text{Battery discharge}
\;\to\; \text{DG} \;\to\; \text{Unmet load.}
\]

Diesel delivery delay is modelled as a geometric process: once an order is
placed, delivery occurs at each subsequent step with a fixed per-step
probability $p$, giving a mean lead time of $1/p$ hours. The geometric
model preserves a Markovian state with only the two pending-order scalars
(flag and quantity), avoiding a countdown vector. Four scenarios are used
in the evaluation campaign: Normal ($p=1/24$, mean 24\,h), Delayed
($p=1/48$, mean 48\,h), Monsoon ($p=1/72$, mean 72\,h), and Extreme
($p=1/336$, mean 336\,h). The learned policies in the primary
comparison---RLInv, TrackB, A5, A6, and A7---were trained under Normal
logistics. All nine policies were then evaluated under Normal, Delayed,
Monsoon, and Extreme conditions.
Each scenario represents a distinct operating regime rather than an
arbitrary parameter value (Table~\ref{tab:scenario_meaning}).
The complete experimental protocol is presented in
Chapter~\ref{chap:experimental-execution}, including the use of ten
sites, ten independent training seeds, and all evaluation scenarios.

\begin{table}[htb]
\centering
\small
\caption{Operational interpretation of the four evaluation scenarios.}
\label{tab:scenario_meaning}
\begin{tabular}{lp{7.5cm}}
\toprule
\textbf{Scenario} & \textbf{Represents} \\
\midrule
Normal   & Routine logistics with nominal supplier performance \\
Delayed  & Moderate but recurring delivery disruption \\
Monsoon  & Seasonal degradation with prolonged uncertainty \\
Extreme  & Severe logistics stress approaching worst-case operating
           conditions \\
\bottomrule
\end{tabular}
\end{table}

Several candidate lead-time parameterisations were explored during
simulator calibration. The four scenarios span routine operation through
severe logistics disruption. Intermediate lead-time settings explored
during calibration produced similar trends and were omitted to keep the
final design interpretable and non-redundant.

\subsection{Safety handling}

Safety is enforced through hard action masking applied at every step. Two
rules are checked before the policy samples an action.

\begin{enumerate}
  \item \textbf{DG masking:} all actions with $\DG_t = 1$ are disabled if
  inventory is below the minimum startup threshold, preventing forced
  mid-run shutdowns.
  \item \textbf{Order masking:} all non-zero order actions are disabled if
  a delivery is already pending, enforcing the single-outstanding-order
  constraint. Order quantities exceeding remaining tank headroom are also
  clipped.
\end{enumerate}

\texttt{MaskablePPO} applies the mask by setting infeasible logits to
$-\infty$ before the softmax; the no-operation action is always feasible
as a safe fallback. Post-step physical clipping (SoC and inventory to
their bounds) provides a numerical backstop and contributes a small
penalty term to the reward (Section~\ref{sec:reward_design}). No adaptive
Lagrangian update was implemented. No hard violations were observed for
the masked policies (verified in Chapter~\ref{chap:results}); the
ablation results there are consistent with an architecture-dependent
benefit from masking under stressed logistics when replenishment is
externally managed, rather than masking acting primarily to eliminate
constraint violations. The A6--TrackB comparison also differs in
observation design (Section~\ref{sec:comparison_policies}), so this gap should not
be attributed to masking alone; the cleaner masking comparison is
RLInv vs.\ A7, which found no significant mean-EENS effect.

\subsection{Agent architecture and training}

Figure~\ref{fig:rlinv_loop} shows the per-step decision loop RLInv
executes at every timestep: the policy proposes an action, infeasible
actions are masked before sampling, the environment steps forward, a
reward is computed, and PPO uses that reward to update the policy over
the course of training.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  node distance=1.6cm,
  box/.style={draw, rounded corners, align=center,
              minimum width=3.2cm, minimum height=0.8cm, font=\small},
  >=Stealth
]
\node[box] (order)   {Policy proposes\\action};
\node[box, right=of order]   (mask)  {Action masking};
\node[box, right=of mask]    (env)   {Environment step};
\node[box, below=of env]     (reward) {Reward computed};
\node[box, left=of reward]   (ppo)    {PPO update};
\node[box, left=of ppo]      (policy) {Policy updated};

\draw[->] (order)  -- (mask);
\draw[->] (mask)   -- (env);
\draw[->] (env)    -- (reward);
\draw[->] (reward) -- (ppo);
\draw[->] (ppo)    -- (policy);
\draw[->] (policy.north) -- (order.south);
\end{tikzpicture}
\caption{RLInv's per-step decision loop. The mask is applied before
sampling, so infeasible actions have zero probability rather than being
sampled and penalised after the fact; PPO's policy update closes the
loop over the course of training.}
\label{fig:rlinv_loop}
\end{figure}

RLInv uses \texttt{MaskablePPO} with a flat MLP $[256,\,256]$ policy.
Training used four parallel \texttt{SubprocVecEnv} environments for
400{,}000 timesteps per seed. Ten independent training seeds were used
to reduce sensitivity to stochastic policy optimisation and to enable
paired statistical comparisons across identical site--scenario
combinations: (42, 123, 777, 7, 13, 21, 99, 314, 500, 999). The first
three (42, 123, 777) were retained from Phase~2 and the
remaining seven were added to increase statistical power for the
seed-level paired comparisons reported in Chapter~\ref{chap:results}. A custom evaluation callback synchronised
\texttt{VecNormalize} statistics between the training and evaluation
environments before each validation pass, preventing observation-scale
mismatch. Training inventory was randomised over $[0.20,\,0.90]\,C^{\Tank}$
so that the policy learns to manage from any starting inventory level;
evaluation randomises initial inventory over $[0.3,\,0.9]\,C^{\Tank}$.

A convergence check extended training to 560{,}000 steps on three
representative sites (1, 5, and~7) using seed~42. Sites~1 and~7 showed no
change in evaluation reward beyond 400{,}000 steps, indicating early
convergence. Site~5 exhibited high reward variance across independent
runs regardless of step count, consistent with its classification as a
grid-constrained site. The 400{,}000-step checkpoint was therefore
retained for all sites and seeds.

Table~\ref{tab:hyperparams} lists all training hyperparameters.

\begin{table}[htb]
\centering
\small
\caption{Training hyperparameters for RLInv and the TrackB dispatch agent.}
\label{tab:hyperparams}
\begin{tabular}{p{3.8cm}p{4.8cm}p{4.4cm}}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Rationale} \\
\midrule
Algorithm              & MaskablePPO (\texttt{sb3-contrib}) & Hard feasibility masking \\
Policy network         & Flat MLP [256,\,256], ReLU         & CPU-feasible capacity \\
Total timesteps        & 400{,}000 per seed                 & Fixed training budget used consistently across all runs \\
Learning rate          & $3\times10^{-4}$ (linear decay)    & PPO standard \\
Clip ratio $\epsilon$  & 0.2                                & PPO standard \\
Entropy coefficient    & 0.01                               & Encourage exploration \\
Discount $\gamma$      & 0.99                               & Effective horizon $\approx$\,100\,h; see Chapter~\ref{chap:results} for a discount-factor sensitivity check against $\gamma=0.995$ \\
GAE $\lambda$          & 0.95                               & Bias--variance balance \\
$n_{\text{steps}}$     & 2048                               & Rollout buffer length \\
Batch size             & 256                                & CPU-friendly mini-batch \\
$n_{\text{epochs}}$    & 10                                 & PPO inner loop \\
Parallel environments  & 4 (\texttt{SubprocVecEnv})         & CPU parallelism; reduced from 8 due to OS process handle limits \\
Seeds                  & 42, 123, 777, 7, 13, 21, 99, 314, 500, 999 & Mean $\pm$ 95\% CI over 10 seeds reported throughout Chapter~\ref{chap:results} \\
\bottomrule
\end{tabular}
\end{table}


% ============================================================
\section{Reward Design}
\label{sec:reward_design}
% ============================================================

The reward is a shaped, normalised scalar designed to penalise
unreliability far more heavily than fuel cost:
\[
r_t =
-\alpha \dfrac{E_t^{\DG}}{\bar{D}}
-\beta  \dfrac{E_t^{\Grid}}{\bar{D}}
-\lambda \dfrac{U_t}{\bar{D}}
-\gamma_r\,\mathbb{I}\!\bigl[U_t>0 \wedge \Inv_t=0\bigr]
-\mu\,V_t,
\]
where $E_t^{\DG}$ is diesel energy consumed, $E_t^{\Grid}$ is grid energy
drawn, $U_t$ is unmet load energy (kWh), $\mathbb{I}[\cdot]$ is a cliff
penalty triggered by a stockout coinciding with an active load shortfall,
and $V_t$ indicates a physical-clipping event (SoC or inventory bound
violation caught by the post-step numerical backstop). Dividing all
energy terms by the site mean hourly load $\bar{D}$ normalises the signal
to a comparable scale across sites of different size. The implemented
reward is a shaped approximation of the economic cost objective defined
in Chapter~\ref{chap:problem}; monetary cost components such as ordering
cost, holding cost, and carbon cost are not explicitly represented.

\textbf{Note on notation.} The stockout-cliff coefficient $\gamma_r$ is
unrelated to the PPO discount factor $\gamma$ used elsewhere in this
chapter; the subscript distinguishes them.

\subsection{Coefficient values and justification}

The default coefficients are $\alpha=1.0$, $\beta=0.4$, $\lambda=100$,
$\gamma_r=200$, $\mu=20$. Each is chosen for a specific, stated reason
rather than by unconstrained tuning:

\begin{itemize}
\item \textbf{$\lambda = 100$} (unmet-load penalty) assigns unmet energy
a substantially larger penalty than diesel use, reflecting that one
kWh of unserved telecom load is far more costly operationally than one
kWh of diesel-generated energy at the most constrained site in the
dataset (site~5). This large ratio is the mechanism by which the reward
prioritises reliability over fuel economy.
\item \textbf{$\gamma_r = 200$} (stockout cliff) is set to exactly twice
$\lambda$, reflecting that a stockout coinciding with unmet load is a
qualitatively different failure mode from unmet load alone (a hard
reliability failure, not merely a cost), and should be penalised more
severely rather than merely proportionally.
\item \textbf{$\beta = 0.4$} (grid-cost weight) reflects that grid
electricity is cheaper than diesel per unit energy in typical Indian
tariff structures, but is not free; an intermediate weight between zero
and $\alpha$ avoids the policy treating grid draw as costless.
\item \textbf{$\mu = 20$} (physical-violation penalty) is deliberately
small: it is a numerical backstop for the rare event that hard masking
still allows a boundary violation (e.g.\ through clipping), not a primary
control lever. Its sensitivity is tested directly in
Chapter~\ref{chap:results}, where broad coefficient ranges are evaluated
across all principal reward parameters, with full values reported there.
\end{itemize}

The selected coefficients encode engineering design preferences rather
than uniquely optimal values, guided by the engineering rationale
discussed below and by prior work
\citep{hu2024performance}. Chapter~\ref{chap:results} evaluates RLInv's
own performance across broad coefficient ranges, including a 20-fold
range for $\lambda$, to assess the sensitivity of RLInv's performance to
these choices.


% ============================================================
\section{Comparison Policies}
\label{sec:comparison_policies}
% ============================================================

Six primary policies and three targeted ablations are evaluated.
Table~\ref{tab:ablation-design} maps every policy to the architectural
question it addresses; Figure~\ref{fig:rlinv_trackb} contrasts the two
principal control architectures directly.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners=4pt, align=center,
              minimum width=3.2cm, minimum height=1.0cm, font=\small},
  arrow/.style={->, thick},
  label/.style={font=\small\bfseries}
]
\node[label] at (3.4, 0.85) {Unified Control (RLInv)};
\node[box] (state1) at (0, 0)
  {System State\\{\scriptsize SoC, Inv, PV, Load, Grid}};
\node[box] (agent1) at (4.2, 0)
  {RLInv Agent\\{\scriptsize MaskablePPO}};
\node[box] (acts1) at (8.4, 0)
  {Actions\\{\scriptsize DG ON/OFF}\\{\scriptsize Diesel Order}};
\draw[arrow] (state1) -- (agent1);
\draw[arrow] (agent1) -- (acts1);
\draw[dashed, gray] (-1.6, -1.05) -- (10.0, -1.05);

\node[label] at (3.4, -1.85) {Decomposed Control (TrackB)};
\node[box] (state2) at (0, -2.5)
  {System State\\{\scriptsize SoC, Inv, PV, Load, Grid}};
\node[box] (ss)       at (4.2, -2.0) {$(s,S)$ Inventory Rule};
\node[box] (order2)   at (8.4, -2.0) {Order Decision};
\node[box] (dispatch) at (4.2, -3.4) {PPO Dispatch Agent};
\node[box] (dg2)      at (8.4, -3.4) {DG Decision};
\draw[arrow] (state2.east) -- ++(0.4,0) |- (ss.west);
\draw[arrow] (state2.east) -- ++(0.4,0) |- (dispatch.west);
\draw[arrow] (ss)       -- (order2);
\draw[arrow] (dispatch) -- (dg2);
\end{tikzpicture}
\caption{Structural comparison of the two principal control
architectures. RLInv jointly optimises generator dispatch and diesel
ordering within a single reinforcement-learning policy. TrackB separates
the two decisions: a classical $(s,S)$ rule handles ordering while a PPO
agent handles dispatch.}
\label{fig:rlinv_trackb}
\end{figure}

\textbf{RLInv} is the proposed unified controller: \texttt{MaskablePPO}
learns to select generator commitment and diesel order quantity jointly
at each step, with all masking rules active.

\textbf{TrackB} is the decomposed-control comparison: a calibrated $(s,S)$
inventory rule determines ordering, while a PPO dispatch agent controls
the generator using the same hidden-layer architecture ($[256,\,256]$
MLP) as RLInv. The dispatch agent's input is a reduced 6-dimensional
observation (state-of-charge, PV, load, grid, and the two cyclic
hour-of-day features only; inventory-related dimensions are withheld
since ordering is handled externally), so "identical network" refers to
matched hidden-layer capacity, not an identical input layer; this keeps
the comparison focused on control structure rather than a confound from
network size.

\begin{quote}
\textbf{Implementation Note.} The reorder point covers estimated mean
consumption over the expected delivery lead time, subject to lower and
upper clipping limits:
$s_{\mathrm{norm}} = \mathrm{clip}(k_s \cdot \bar{D} \cdot \mathbb{E}[L]
\,/\, C^{\Tank},\, 0.05,\,0.95)$ with safety multiplier $k_s=1$, and
order-up-to level $S_{\mathrm{norm}} = \mathrm{clip}(0.85,\,
s_{\mathrm{norm}}+0.10,\,0.95)$. This calibration is shared by B1,
TrackB, A6, MPC, and Oracle-MPC.
\end{quote}

\textbf{MPC} pairs the same $(s,S)$ ordering rule with a model-predictive
dispatch controller: a receding-horizon MILP (solved via HiGHS,
\texttt{scipy.optimize.milp}) re-optimised at every step over a
24-hour planning horizon, using a persistence forecast (the current
observation repeated across the horizon). A 24-hour horizon captures the
daily solar and load cycle while requiring only 24 binary dispatch
variables, keeping the MILP computationally small. MPC's battery--DG
co-optimisation within the MILP is a mild optimistic bias relative to
RLInv's greedy priority-rule battery dispatch, since the solver can plan
battery and generator jointly where the simulator's own dispatch cannot;
this is a documented, deliberate advantage given to the baseline. Under
Extreme logistics, the expected delivery time lies far beyond the
24-hour planning horizon (mean lead time 336\,h), limiting MPC's ability
to account for replenishment within its current optimisation window ---
the central structural limitation this comparison is designed to expose. A
third variant using a learned 24-hour forecasting module in place of the
persistence forecast was also implemented and evaluated during
development; it is not part of the primary comparison, as it did not
alter the qualitative conclusions drawn from the persistence-forecast
MPC baseline.

\textbf{Oracle-MPC} uses the realised future load, solar-generation and
grid-availability traces over the remaining evaluation horizon (perfect
information) rather than a persistence estimate, and its planning
horizon extends to the full 360-step evaluation episode rather than 24
hours. Diesel ordering remains governed by the same $(s,S)$ rule, and
delivery timing remains stochastic. It is not a
deployable policy; it exists to answer a specific diagnostic question: if
the gap between RLInv and TrackB/MPC were explained purely by better
forecasting or better dispatch optimisation, RLInv should not be able to
match or approach a policy with \emph{perfect} forecasting and dispatch.
Since RLInv's ordering is still learned rather than fixed, a persisting
RLInv--Oracle-MPC gap in either direction is informative about whether
the remaining advantage is attributable to ordering rather than dispatch
quality (Chapter~\ref{chap:results}).

\textbf{B0} is a rule-based heuristic: a fixed-threshold inventory
policy using uncalibrated inventory thresholds, paired with threshold
dispatch, requiring no training. \textbf{B1} replaces B0's fixed
thresholds with the same site-calibrated $(s,S)$ inventory policy used
by TrackB, MPC, and Oracle-MPC; B0 and B1 share the same dispatch
heuristic, differing only in the ordering rule. Concisely: B0 uses fixed
inventory thresholds, B1 uses a site-calibrated $(s,S)$ inventory
policy --- isolating the value of calibrated ordering specifically.

\textbf{A7} is vanilla PPO trained without action masking, isolating the
contribution of the masking safety mechanism.

\textbf{A5} removes five logistics/inventory-related channels via a
\texttt{NoInvObsWrapper}: inventory level ($\Inv_t$), pending-delivery
flag, pending quantity, elapsed time since ordering ($\tau_t$), and the
remaining delivery-time channel ($R_t$); the remaining six dimensions
(SoC, PV, load, grid, and the two cyclic hour-of-day features) are
unchanged. This tests the combined contribution of inventory,
pending-delivery, and ordering-history information; because five
channels are removed together, their individual effects are not
isolated.

\textbf{A6} provides a cleaner ordering ablation while acknowledging
that the dispatch policy is re-learned under a reduced action space,
rather than a perfectly controlled isolation of the ordering decision
alone. In
this variant the RL agent controls only generator commitment under a
reduced \texttt{Discrete(2)} action space, while diesel ordering is
handled by the same calibrated $(s,S)$ rule used in TrackB, MPC, and
Oracle-MPC. The observation space, reward function, and environment
dynamics remain identical to RLInv. Because the dispatch policy is
re-learned under the smaller action space, A6 provides a cleaner ordering
ablation than TrackB, which confounds ordering mechanism and dispatch
architecture simultaneously; both comparisons are reported in
Chapter~\ref{chap:results} for this reason.

\begin{table}[htb]
\centering
\small
\caption{Policies evaluated in the experimental campaign, the question
each answers, and the classical $(s,S)$ ordering rule shared across four
of the six comparators.}
\label{tab:ablation-design}
\begin{tabularx}{\textwidth}{
    >{\raggedright\arraybackslash}p{1.4cm}
    >{\raggedright\arraybackslash}p{2.15cm}
    >{\raggedright\arraybackslash}p{2.25cm}
    >{\raggedright\arraybackslash}p{2.0cm}
    >{\raggedright\arraybackslash}X
}
\toprule
\textbf{Label} &
\textbf{Ordering} &
\textbf{Dispatch} &
\textbf{Forecast} &
\textbf{Primary question} \\
\midrule
B0 &
Fixed threshold (uncalibrated) &
Threshold rule &
--- &
Baseline anchor \\
B1 &
$(s,S)$ rule &
Threshold rule &
--- &
Value of intelligent ordering \\
MPC &
$(s,S)$ rule &
MILP, $H{=}24$ &
Persistence &
Does optimal-but-myopic dispatch close the gap? \\
Oracle-MPC &
$(s,S)$ rule &
MILP, $H{=}360$ &
Perfect &
Optimistic reference with perfect future load, PV, and grid availability \\
TrackB &
$(s,S)$ rule &
Learned (PPO) &
--- &
Joint vs.\ decomposed (H1, confounded) \\
A6 &
$(s,S)$ rule &
Learned (PPO, Discrete(2)) &
--- &
Joint vs.\ decomposed (H1, cleaner) \\
A7 &
Learned (joint) &
Learned (PPO, no mask) &
--- &
Action masking (H2) \\
A5 &
Learned (joint, no inv.\ obs.) &
Learned (PPO) &
--- &
Inventory observability (H3) \\
RLInv &
Learned (joint) &
Learned (PPO) &
--- &
Proposed system \\
\bottomrule
\end{tabularx}
\end{table}


% ============================================================
\section{Experimental Protocol}
\label{sec:evaluation_protocol}
% ============================================================
This section specifies the experimental protocol required to reproduce
the evaluation, including training, testing, ablation, robustness, and
statistical comparison procedures. The quantitative results obtained
using this protocol are presented separately in
Chapter~\ref{chap:results}.

Figure~\ref{fig:eval_flow} summarises the overall protocol: policies are
trained, then evaluated on the held-out test window, with ablations and
sensitivity checks run against the same evaluation grid before the
paired statistical analysis is applied.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  node distance=1.2cm,
  box/.style={draw, rounded corners, align=center,
              minimum width=4.4cm, minimum height=0.8cm, font=\small}
]
\node[box] (train)      {Training (10 sites $\times$ 10 seeds)};
\node[box, below=of train]      (evaluate)  {Evaluation (4 scenarios, 10 episodes/seed)};
\node[box, below=of evaluate]   (ablations) {Ablations (A5, A6, A7 vs.\ RLInv)};
\node[box, below=of ablations]  (sensitivity) {Sensitivity checks (reward, tank,
                                                distribution, grid)};
\node[box, below=of sensitivity] (statistics) {Paired statistics (Chapter~\ref{chap:results})};

\draw[->] (train)      -- (evaluate);
\draw[->] (evaluate)   -- (ablations);
\draw[->] (ablations)  -- (sensitivity);
\draw[->] (sensitivity)-- (statistics);
\end{tikzpicture}
\caption{Overall evaluation protocol, from training through to the
paired statistical analysis reported in Chapter~\ref{chap:results}.}
\label{fig:eval_flow}
\end{figure}

All nine policies are evaluated on the 15-day held-out test window (days
46--60) of each of the ten ITU/Zindi sites, under all four delivery
logistics scenarios (Normal, Delayed, Monsoon, Extreme; defined in
Section~\ref{sec:env_impl}), with 10 episodes per seed and initial
inventory randomised uniformly over $[0.3,\,0.9]\,C^{\Tank}$. This is a
full factorial design: every policy is evaluated under every scenario at
every site, rather than the scenario coverage being policy-specific.
Table~\ref{tab:protocol_summary} summarises the evaluation grid at a
glance.

\begin{table}[htb]
\centering
\small
\caption{Experimental protocol at a glance.}
\label{tab:protocol_summary}
\begin{tabular}{lp{10.5cm}}
\toprule
\textbf{Item} & \textbf{Value} \\
\midrule
Sites                & 10 (all ITU/Zindi sites) \\
Training seeds       & 10 (42, 123, 777, 7, 13, 21, 99, 314, 500, 999) \\
Evaluation episodes  & 10 per seed \\
Scenarios            & 4 (Normal, Delayed, Monsoon, Extreme) \\
Policies evaluated   & 9 (B0, B1, MPC, Oracle-MPC, TrackB, A5, A6, A7, RLInv) \\
Primary metric       & EENS (kWh per episode) \\
Secondary metrics    & Diesel consumption, stockout events, orders placed, inventory trajectory \\
Statistical unit     & Seed-level mean, paired across policies \\
\bottomrule
\end{tabular}
\end{table}

Statistical comparisons use the seed-level mean as the unit of analysis
($n=10$ seeds per condition, or $n=100$ when pooling across all 10 sites
$\times$ 10 seeds for the primary comparison). Hypothesis comparisons
(H1, H2, H3) are assessed with a paired $t$-test on within-seed
differences, since the same 10 seeds are evaluated under every policy
(seed 42 under RLInv is matched against seed 42 under the comparator, not
treated as an independent sample), with effect size reported as Cohen's
$d_z$ (mean paired difference divided by the standard deviation of the
paired differences). Because every policy was evaluated on identical
site, scenario, and seed combinations, paired analysis provides greater
statistical power than an independent-samples test by removing
between-instance variability that is common to both policies being
compared. A subset of three sites (site2, site5, site7) is
additionally used for finer-grained robustness checks in
Chapter~\ref{chap:results} where a smaller, more tractable evaluation
grid is needed (e.g.\ reward and tank-capacity sensitivity sweeps); this
three-site subset was selected for load/solar/outage diversity and is
distinct from the site difficulty classification of
Chapter~\ref{chap:datasets-setup}, where only site2 and site5 are
classified Hard.

Full numerical results, including the primary comparison, the three
hypothesis tests, behavioural analysis, and all robustness checks
(reward weighting, discount factor, tank capacity, delivery-distribution
form, stochastic grid outages, and the RLInv-DA exploratory extension),
are reported in Chapter~\ref{chap:results}. All reported
experiments used ten evaluation episodes per training seed.


% ============================================================
\section{Implementation Deviations from the Formal Model}
\label{sec:impl_deviations}
% ============================================================

\subsection{Deviations from the formal CMDP}
\label{sec:deviations_table}

Although the implementation follows the CMDP formulation of
Chapter~\ref{chap:problem} closely, several practical simplifications
were adopted to enable stable training within the available CPU budget.
Table~\ref{tab:deviations} consolidates these points; each deviation was
adopted for a specific practical reason and does not invalidate the
empirical study, but defines the scope within which results should be
interpreted.

\begin{table}[htb]
\centering
\small
\caption{Deviations between the formal CMDP (Chapter~\ref{chap:problem})
and the implemented system.}
\label{tab:deviations}
\begin{tabular}{p{2.8cm}p{4.6cm}p{4.6cm}}
\toprule
\textbf{Aspect} & \textbf{Formal model} & \textbf{Implementation} \\
\midrule
Battery control
  & $P_t^{\Bat}$ included in CMDP action space
  & Fixed energy-priority rule; not an RL decision \\[4pt]
Action space
  & Hybrid continuous $+$ discrete; order quantity a fixed litre value
  & Discrete 6-action encoding; order quantity expressed as \{0, 30\%,
    60\%\} of tank capacity rather than a fixed litre amount \\[4pt]
Pipeline representation
  & Single scalar $\Pipe_t$ (quantity in transit)
  & Represented as a flag $+$ quantity pair (two observation
    dimensions) rather than a single scalar; mathematically
    equivalent since at most one order is ever in transit \\[4pt]
Discount factor
  & $\gamma = 0.995$ (effective horizon $\approx$\,200\,h)
  & $\gamma = 0.99$ (effective horizon $\approx$\,100\,h; sensitivity to
    this choice is tested directly in Chapter~\ref{chap:results}) \\[4pt]
Reward
  & Economic cost decomposition (monetary)
  & Shaped normalised scalar (Section~\ref{sec:reward_design}) \\[4pt]
Safety mechanisms
  & Masking $+$ recovery $+$ Lagrangian
  & Masking $+$ physical clipping only \\[4pt]
Neural architecture
  & Modular domain-specific encoders
  & Flat MLP $[256,\,256]$, ReLU \\[4pt]
Training budget
  & 576{,}000 steps (script default)
  & 400{,}000 steps (final run); convergence check extended training to
    560{,}000 steps confirming plateau \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Diesel supply-chain scope and assumptions}
\label{sec:supply_chain_scope}

This study models the diesel supply chain using the following
assumptions and scope boundaries, consolidated here for clarity, with
the corresponding formulation and implementation details presented in
Chapter~\ref{chap:problem} and Table~\ref{tab:deviations} above.

\textbf{Single vendor, single outstanding order.} The simulator models
one delivery pipeline per site with at most one order in transit at a
time; multi-vendor sourcing, vendor selection, and contractual or
pricing considerations are not represented. This keeps the pending-order
state a two-scalar quantity (Section~\ref{sec:env_impl}) rather than
requiring a general queue.

\textbf{Delivery delay as a stochastic process, not a routing problem.}
Delay is modelled as a per-site geometric (or, for robustness,
lognormal/Weibull) random process calibrated to a mean lead time, not as
the output of an explicit vehicle-routing, road-network, or
depot-inventory simulation. This is a deliberate abstraction: it
represents the \emph{effect} of upstream logistics uncertainty on a
single tower's replenishment decisions without requiring a full
logistics network model, which would require a fundamentally different
problem formulation centred on inventory-routing optimisation rather
than single-site inventory-aware control.

\textbf{Single-tower scope.} Each site is optimised independently; no
shared supplier capacity, delivery-slot competition, or fleet-level
coordination across towers is modelled (Chapter~\ref{chap:conclusion}
discusses a multi-agent extension as future work).

\textbf{No real vendor delivery data.} The ITU/Zindi dataset does not
record delivery events, so every delay distribution used in this thesis
--- including the lognormal and Weibull robustness checks
(Chapter~\ref{chap:results}) --- is a parametric modelling choice, not
one fitted to observed logistics records. This is stated here as a
scope boundary, not revisited as a limitation each time the delay model
is used elsewhere in this thesis.

The RLInv-DA extension (Chapter~\ref{chap:results}) addresses a related
but distinct question --- whether a persistent, autocorrelated supplier
condition (rather than a fully independent-order abstraction) is
worth tracking explicitly --- and its own scope is documented alongside
that result rather than repeated here.